%        File: paper1.tex
%     Created: Wed Oct 12 02:00 AM 2011 E
% Last Change: Wed Oct 12 02:00 AM 2011 E
\documentclass[12pt,letterpaper]{article}
\usepackage{setspace}
\usepackage[margin=1.1in]{geometry}
\doublespacing
\begin{document}
\title{Youth Indifference in American Politics: Causes, Consequences,
and Solutions}
\author{Justin Cullen}
\date{\today}
\maketitle

\section{Introduction} % (fold)
\label{sec:Introduction}
It's been a common trend throughout American history for young
Americans to be less involved than their older counterparts in
political issues.  In recent decades however, this gap in the
participation of younger Americans relative to old has noticeably
increased. In polls taken during the 1990s, American's aged 18-29 were
more than 20\% less likely to look forward to reading a newspaper, or
watching news media than those above 30. Data from the 2010 census
shows that American's between 18-29 make up nearly 30\% of the total
American population, so this data suggests that nearly a third of
American citizens show signs of disinterest in political affairs.  

If we accept that civil engagement of citizens is a fundamental
necessity for the functioning of our democratic system, this is a
disturbing trend indeed.  The purpose of this paper is threefold:
First, I establish some of the most important reasons that
indifference to political issues is becoming more widespread among
young Americans.  Next, I argue that if left unchecked, this trend
could have serious consequences for Americas' future.  Finally, I
offer some steps that can be taken now to help to reverse this trend
and avert these consequences.
% section Introduction (end)

\section{Why Indifference?} % (fold) 
\label{sec:Why Indifference?}

\subsection{Lack of impact on outcome of political processes} There
are numerous reasons why a young American might refrain from taking an
active interest in following the political events surrounding him.
The first of these stems from the perceived scale of the political
landscape itself.  Many young Americans feel that their individual
contribution to the outcome of political processes is negligible, and
this can cause them to refrain from participating in it at all.  Often
young people believe that political decisions are simply made by older
politicians, and regardless of what they do, the outcome will be the
same anyways.  This feeling of inevitably is common particularly among
Americans who are just coming into adulthood and trying to find their
place in the world. ``My actions have no impact, so why do anything?''
is unfortunately a surprisingly feeling among young voters.

\subsection{Distrust of the Political Process} In addition to being
intimidated by the scale on which political decisions are made, many
young Americans are dissuaded from participating in politics by a
distrust in the integrity of the process itself. In interviews
conducted of a sample of American college students, there was an
overwhelming amount of cynicism regarding the character and
motivations of politicians.  Politicians were often perceived by the
interviewees as ``dishonest'',``corrupt'',``insincere'', or
``self-serving''.  This is an far stronger motivation to refrain from
participating in the political process than the previous example of
size. If someone doesn't have faith that the political outcomes are
determined fairly by the majority will of the people, they have even
less incentive to be engaged in the process than someone who is
deterred by the scale of the system alone.  They might feel that not
only is their individual vote insignificant, but that even if that
manage to gather a sizeable number of supporters, political outcomes
could still be unchanged.

\subsection{Lack of Partisanship} Another factor that contributes to
the young population of America's lack of participation in political
issues is that of partisanship.  Results of polls consistently show a
positive correlation between strong attachment to a political party
and general political involvement.  People with strong partisan views
will naturally have stronger preferences for particular policies and
politicians and thus will be more likely to vote and keep up with
current events. Surveys conducted in the late 1990s found that only
21\% of Americans under 30 considered themselves strong partisans
versus 46\% for Americans above 30. Older Americans are much more
likely to have strong partisan views than young Americans.  This makes
sense - younger Americans simply have had less time to develop rigid
opinions about the world around them and align themselves with
particular causes or parties.  

\subsection{Ineffective Media Reach to Youth} I would argue that this
final reason for youth apathy toward politics is one of the most
important, but it is also the one most easily remedied.  It's
impossible to ignore the huge difference in the amount that younger
Americans follow news media compared to their older counterparts.
Polls show that while 76\% of Americans over age 30 reported reading a
newspaper the day before they were polled, only 31\% of those below
age 30 had. Similarly, in the same poll, 86\% of the over-30 group
reported watching a TV news show the day before they were interviewed,
compared with only 45\% of those under 30.  

We must accept the possibility that current primary methods of
communicating current events in politics - namely news TV and paper
newspapers - are simply ineffectively doing the job when it comes to
the younger contingent of voting aged Americans.  One psychological
explanation for this is that children growing up typically associate
reading newspapers and watching the news as something their parents do
- something boring for ``old people''.  Even as children come into
adulthood and reach the age where they themselves should be voting,
this association often remains, causing them to avoid these activities
and thus remained uninformed about current events. Later, I will argue
that the rise of the internet as a medium for communication gives us a
unique opportunity to reach younger audiences without this attached
stigma.  First however, I will discuss the ``why'' of doing this, by
examining some potential consequences of leaving this trend to
continue unchecked.
% section Why Indifference? (end)

\section{Consequences of Youth Indifference} % (fold)
\label{sec:Consequences of Youth Indifference}
On a basic level, the consequences of young Americans' indifference to
public affairs all stem from an assumption that participation in the
political system is a fundamental requirement for being a responsible
citizen in a democratic society.  Brady, Schlozman, Verba convincingly
argue the validity of this assumption in a paper titled ``Civic
Participation and the Equality Problem''.  In this paper, the authors
point out that political inactivity is not uniformly distributed
across the population.  In fact, it is quite the opposite as we have
already seen here in regard to age.  In their paper, they focus on
imbalances in political participation between social classes. They
point out the fact that Americans who are less financially well off
are less likely to be politically involved.  This means that this
broad section of the population is under-represented, and their
viewpoints and concerns are likely to receive less attention and
representation than is warranted.  This is in direct opposition to one
of the fundamental tenets of democratic society - namely that the
needs and preferences of all citizens should be equally represented in
government.

\subsection{Under-representation of Youth Needs}
I would argue that this analysis applies equally to the case of
age-bias in political involvement.  In the same way as above, a
tendency for political indifference in young voters inadvertently
causes their needs and preferences to be under-represented in
government.  There are many policy decisions which have significantly
more impact on the younger citizens than on those who are middle aged
and above.  Examples of these include healthcare, social security, and
the environment. Over a third of Americans are under age 30 and have
an enormous stake in these types of issues, yet the abstaining from
civic engagement causes their concerns and needs relating to these
issues to be under-emphasized.

\subsection{Impact on Personal Development} A second argument made for
the importance of civic participation in the Brady paper concerns the
development of the individual, rather than the representation of the
groups needs. The Brady paper argues that civic engagement is
paramount in developing many of the desirable characteristics we
associate with responsible citizens.  Among them are independence,
critical thought, respect for others, and a willingness to take on
responsibility. This is of particular concern when applied to the case
of younger Americans. These younger Americans are the future leaders
of the country! If it is indeed true that participation in ones civic
duties are a critical part of learning to become a respectable adult,
then we simply cannot afford to have our youth be uninvolved.  Of
course you might say that it is natural for younger people to be less
interested in politics and that they will ``grow into'' their roles as
responsible citizens with time.  I would argue that this is not a
chance we can afford to take. Amidst a long term trend of increasing
indifference, cannot simply sit back and hope that young Americans'
will ``come around'' in time to lead the country when there are steps
we can take now to combat this trend.
% section Consequences of Youth Indifference (end)

\section{Remedies}
I propose two changes that I feel will help to increase youth interest
in American political affairs.  Of the causes presented in section 1,
I feel that the ineffective use of the media as a communication medium
to young voters is the most tangible, and easiest to address, so I
will focus my attention on that issue. The first change addresses the
stigma associated with the news as something that young people
shouldn't be interested in.  The second change relates to how to more
effectively use the internet as a communication medium to get
political news to young America.

\subsection{De-Stigmatizing the News}
As described in section two, young people often carry the belief that
watching television news or reading newspapers is something for ``old
people'' to do.  This belief is a result of growing up seeing their
parents watch the news every night and likely being bored by it.
This is understandable, as the main news programs primarily target
adults and aren't designed to hold the attention of younger children.
But what if there were television news shows and newspapers written
specifically to interest younger viewers?  In fact, these types of
shows do exist, but they aren't particularly widespread. For
television shows, Nick News and Channel One News are the most popular
examples.  For print media, there aren't any notable national
newspapers specifically targeting a youth audience. I believe that
developing more TV news programs and newspapers with language and
topics crafted to be of interest to younger audiences would do a lot
to engage youth in politics.  

\subsection{Making Better Use of the Internet}
Only in the past decade has internet use become widespread enough to
warrant use as a platform to communicate political information.  As
early as the beginning of the decade, we saw the first websites
dedicated to the campaigns of political candidates.  Today, the
majority of candidates running for public office have websites for
their campaigns, but they are not as effective as they could be in
bringing in younger voters.  

% section section name (end)

\section{Conclusion} % (fold)
\label{sec:Conclusion}
We have seen a broad overview of the topic of political indifference
among young Americans. Although some of the causes for this
indifference discussed earlier are out of our immediate control, many
aren't. Given the importance of having a politically active
population, we need to take steps as a nation to address this problem
and reengage the younger voters of America.  Not only are the young
the future leaders of the America, they are also the demographic that
- when politically informed - is most likely to get up and take action
now. We need to do something to re-involve American youth in politics,
and I believe the two suggestions outlined here are a good place to
start.  Political apathy among the young is not a sentiment we can
afford to spread.

% section Conclusion (end)

\end{document}

